---
title: "Data challenge"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### R shiny app
A shiny app was made to visualize Y chromosome, one can locate exact regions with losses.  
1. Open runShiny.R, change path to the folder with data files, ui.R and server.R files.  
2. Run all scripts in runShiny.R, a window with panel will appear, it may take a few seconds.  
3. Select one sample to continue, a plot showing read depth on y-axis and location on x-axis will appear.  
4. Move cursor on the plot, it will display depth and location information.   

#### Copy number prediction 
Prepare environment, load data, use batch2 as example, same process for batch1.
```{r}
setwd('/DataChallenge/Batch_2')
#savehistory('data-challenge-all.rhistory')

depth = read.csv('Depth.csv', header = T)
interval = read.csv('Interval_order.csv', header = T)
```

  
Visualization. Blue lines are moving average of depth, x-axis as location from chr1 to chrY, y-axis as read depth. Red lines denote the regions for X, Y chromosomes. I did not take into account the chromosomal location for each interval, because not all regions are equally covered. 
```{r, fig.width=15, fig.height=15}
library(TTR)
par(mfrow=c(6,4))
par(mar=c(2.5,2,2,1))
# plot simple moving average for all chromosomes
for(i in 1:24){
    plot(SMA(depth[,i], n=1000), col="blue", main=paste0('sample',i), type='l', xlab='location', ylab='average depth')
    abline(v=c(which(interval$Chrom==23)[1], which(interval$Chrom==24)[1]), col='red' )
}
dev.copy2pdf(file='SMA-plot.pdf', width=10, height=10)
# plot X
for(i in 1:24){
    plot(depth[(which(interval$Chrom==23)[1]):(which(interval$Chrom==24)[1]-1),i], col="black", main=paste0('sample',i), type='l', xlab='location', ylab='average depth', ylim=c(0,2000))
}
dev.copy2pdf(file='raw-x-plot.pdf', width=15, height=15)
# plot Y
for(i in 1:24){
    plot(depth[(which(interval$Chrom==24)[1]):dim(interval)[1],i], col="black", main=paste0('sample',i), type='l', xlab='location', ylab='average depth', ylim=c(0,2500))
}
dev.copy2pdf(file='raw-y-plot.pdf', width=15, height=15)
dev.off()
```

  
PCA also helps to distinguish samples with copy number abnormalities.
```{r, fig.width=6, fig.height=6}
## pca with top 5000 intervals with high variance 
plot.pca <- function(ntop, depth){
    library(matrixStats)
    rv <-  rowVars(as.matrix(depth))
    select <-  order(rv, decreasing = TRUE)[seq_len(ntop)]
    pca <-  prcomp(t(depth[select, ]))
    plot( pca$x[,1], pca$x[,2], xlab= 'PC1', ylab='PC2' )
    text( pca$x[,1], pca$x[,2], seq(1,24), pos= 2 )
}
plot.pca(5000, depth)
dev.copy2pdf(file='pca-top5000.pdf', width=8, height=8)
```

  
Calculate copy number for chromosome X and Y, write output files
```{r}
get.copy <- function(depth){
    # get average read depth for chr 1-22
    avg <-  colMeans(depth[1:(which(interval$Chrom==23)[1]-1),], na.rm=T)
    # get average read depth for X, Y
    avg.x <- colMeans(depth[(which(interval$Chrom==23)[1]):(which(interval$Chrom==24)[1]-1),], na.rm=T)
    avg.y <-  colMeans(depth[(which(interval$Chrom==24)[1]):dim(interval)[1],], na.rm=T)
    # calculate fold change for chr X, Y, divide by 0.5 because chr 1-22 are diploidy
    copy.x <-  round(avg.x/ avg, digit=2)/0.5
    copy.y <-  round(avg.y/ avg, digit=2)/0.5
    copy <-  data.frame(copy.x, copy.y) 
    copy$id <- rownames(copy)
    copy <- copy[,c(3,1,2)]
    return(copy)
}
copy <- get.copy(depth)
write.table(copy, file='estimated-xy-copy.txt', sep='\t', quote=F, row.names = F)
library(knitr)
kable(copy[,2:3], caption='Prediction')
```
  

Improvement:  
Given time, rather than taking X and Y chromosome as a whole, more complex method can be developed to take into account regional loss of Y so as to give more precise copy number prediction. It might also be helpful to remove the extreme values and perform normalization across samples. 
Given .bam and other files, more sophisticated methods, such as bubbletree and MADSEQ from Biodonductor, can be applied to get regional aneuploidy information.

